---
title: "R Notebook"
output:
  pdf_document:
    toc: true
  html_document:
    df_print: paged
  word_document:
    toc: true
---


\pagebreak

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
rm(list=ls())
```


```{r}
library(dplyr)
library(forcats)
library(survey)
library(tidyr)
library(car)
library(haven)
library(survey)
library(tidyr)
library(broom)
library(ggplot2)
library(ggeffects)
library(ggeffects)
library(Hmisc)
library(readxl)

# new 2023 ACS data
undoc23e11 <- read_dta("/Users/estellepan/Desktop/AIC/undocumented_student/undoc_2023_acs_02_27_2025.dta")

# state-level enrollment & employment data 
state_enroll <- read.csv("/Users/estellepan/Desktop/AIC/undocumented_student/state_enroll_long(Sheet1).csv")
```


```{r}
# Filter for undocumented youth aged 18–24 
# who have completed high school (GED or diploma) 
# but have NOT completed a college degree

undoc_youth <- undoc23e11 %>%
  filter(
    undoc2 == 1,                  # Undocumented immigrants (clean, NA-free flag)
    age >= 18 & age <= 24,        # Target age range per study definition
    educd %in% c(                 # Educational attainment codes (2023 ACS only):
      063,  # Regular high school diploma
      064,  # GED or alternative credential
      065,  # Some college, less than 1 year
      071,  # 1+ years of college credit, no degree
      081   # Associate's degree, type not specified
    )
  )

##  People who completed HS and are in college or have some college, exclude people who already have college degrees 
```


```{r}
# dataset excluding people who already have college degrees 
undoc23e1 <- undoc_youth %>%
  left_join(state_enroll, by = c("statefip" = "StateFIP"))
```


```{r}
# remove labelled metadata 
undoc23e1<-zap_labels(undoc23e1)
undoc23e1$state_employ_rate<-zap_formats(undoc23e1$state_employ_rate)
undoc23e1$state_enroll_rate<-zap_formats(undoc23e1$state_enroll_rate)
summary(undoc23e1$state_employ_rate)
summary(undoc23e1$state_enroll_rate)
```

```{r}
# Explore Key Variables 
summary(undoc23e1$age)
table(undoc23e1$undoc)
```

```{r}
colSums(is.na(undoc23e1))
```



# Cleaning variables again and doing some prelim stats before regressions

## Re-making state employ into a factor incase of linearity of logit violations

```{r}
undoc23e1$state_employ23 <- as.numeric(undoc23e1$state_employ_rate)
hist(undoc23e1$state_employ23)
```

```{r}
# Why square it? To capture curvature in the relationship if state_employ_rate has a nonlinear effect on outcome.
undoc23e1$state_emp23_sq<-undoc23e1$state_employ23^2
hist(undoc23e1$state_emp23_sq)
```

```{r categorical emp rate 23}
summary(undoc23e1$state_employ23)
undoc23e1$state_employ23_cat<-cut(undoc23e1$state_employ23,
                           breaks=3,
                           labels=c("1","2","3"))

table(undoc23e1$state_employ23_cat)
```

```{r}
undoc23e1%>%group_by(state_employ23_cat)%>%
  summarise(min=min(state_employ23),
            max=max(state_employ23))
```  


## remaking state enrollment into a factor incase of linearity of logit violations 


```{r doing a squared enroll rate}
undoc23e1$state_enroll23 <- as.numeric(undoc23e1$state_enroll_rate)
hist(undoc23e1$state_enroll23)
```

```{r}
undoc23e1$state_enroll23_sq<-undoc23e1$state_enroll23^2
hist(undoc23e1$state_enroll23_sq)
```


```{r}
undoc23e1$state_enroll23_cat<-cut(undoc23e1$state_enroll23,
                           breaks=3,
                           labels=c("1","2","3"))

table(undoc23e1$state_enroll23_cat)

undoc23e1%>%group_by(state_enroll23_cat)%>%
  summarise(min=min(state_enroll23),
            max=max(state_enroll23))
```


### Household income categories
```{r}
# Use quantile() with cut() to ensure balanced group sizes
undoc23e1$hhincome_cat <- cut(
  undoc23e1$hhincome,
  breaks = quantile(undoc23e1$hhincome, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  labels = c("1", "2", "3"),
  include.lowest = TRUE
)
table(undoc23e1$hhincome_cat)

undoc23e1 %>%
  group_by(hhincome_cat) %>%
  summarise(
    min = min(hhincome, na.rm = TRUE),
    max = max(hhincome, na.rm = TRUE)
  )

```


```{r}
summary(undoc23e1$age)
```


# Logistic Regression Set Up



## setting up the svy object 
```{r making person id variable}
undoc23e1$ids<-undoc23e1$serial+undoc23e1$pernum
```

## final model we didn't use race
```{r race4 and race5}
undoc23e1$race3 <- factor(undoc23e1$race,
                         levels = 1:9,
                         labels = c("White", "Black", "AIAN", "Chinese", "Japanese", 
                                    "API_other", "Other", "TwoRaces", "ThreePlus"))
```

```{r}
#  collapsing "Asian or Pacific Islander (other)" into a broader "other" group
undoc23e1$race4<-fct_recode(undoc23e1$race3,
                           "other"="API_other")
# Sets "Black" as the reference group for modeling in race5
undoc23e1$race5<-relevel(undoc23e1$race4, ref="Black")
# using the original 9-category race3
undoc23e1$race6<-relevel(undoc23e1$race3, ref="Black")
```

```{r setting up interactions for box tidewell}
# helps detect or model non-linear effects in logit models for continuous predictors
undoc23e1$box_enroll23<-undoc23e1$state_enroll23*log(undoc23e1$state_enroll23)
undoc23e1$box_employ23<-undoc23e1$state_employ23*log(undoc23e1$state_employ23)
```


```{r}
#Create employ Variable
undoc23e1$employ <- ifelse(undoc23e1$empstat == 1, 1, 0)
```

```{r}
# Define state FIPS codes for in-state tuition access
# WA, OR, CA, NV, UT, AZ, CO, NM, NE, KS, OK, TX, HI, MN, IL, KY, VA, FL, NY, VT, MA, CT, RI, NJ, MD, DC
isrt_states <- c(53, 41, 6, 32, 49, 4, 8, 35, 31, 20, 40, 48, 15,
                        27, 17, 21, 51, 12, 36, 50, 25, 9, 44, 34, 24, 11)

# Create binary indicator variable
undoc23e1$isrt <- ifelse(
  undoc23e1$statefip %in% isrt_states, 1, 0
)
```

```{r}
# Define state FIPS codes for states that allow driver's licenses for undocumented immigrants
drive_lic_states <- c(6, 8, 9, 10, 11, 15, 17, 24, 25, 27, 32, 34, 35, 36, 41, 44, 49, 50, 51, 53)

# Create binary indicator variable
undoc23e1$driveLic <- ifelse(undoc23e1$statefip %in% drive_lic_states, 1, 0)
```


# combined Central America, Caribbean, and South America as 1 and the rest as 0.  


```{r}
# Used `dpl` (summary birthplace variable) instead of `dpld` because:
# 1. `dpl` has full and consistent coverage across the ACS 2023 sample.
# 2. `dpld` is more granular but includes many missing or unavailable categories in the current dataset.
# 3. `dpl` captures all major Latinx regional categories needed (e.g., Central America, Caribbean, South America, excluding mexico).
# 4. Using `dpl` maintains consistency with prior studies' region-level groupings.
undoc23e1$birthplace1 <- case_when(
  undoc23e1$bpl %in% c(210, 250, 260, 299, 300) ~ "Cam Sam",  # Central America, Caribbean, South America
  TRUE ~ "Other"
)
```

```{r}
# subset for lat but excluding Mexico , use this for further regression 
lat_sub1 <- subset(undoc23e1, birthplace1=="Cam Sam")
```

```{r states in model}
undoc23e1%>%group_by(state)%>%
  summarise(n=sum(perwt))%>%
  print(n=50)
```






```{r}
save(undoc23e1, file="final_undoc23e.RData")
```


```{r}
# svydesign 
# creates a survey-weighted design object using perwt (person weight).
undoc_data1<-svydesign(id=~ids, 
                         weights=~perwt,
                       data=undoc23e1) 
```

```{r}

lat_sub1<-svydesign(id=~ids, 
                         weights=~perwt,
                       data=lat_sub1) 

```


# Running Regressions 

```{r}
# check following variables used in the prior regression model are not found in the current dataset:

model_vars <- c("attending_college", "isrt", "driveLic", "sex", "age",
                "birthplace1", "daca_imm", "employ", "hhincome_cat",
                "state_enroll23", "state_employ23", "box_enroll23", "box_employ23")

missing_vars <- model_vars[!(model_vars %in% names(undoc_data1$variables))]
missing_vars
```

### Box Tidewell for state enrollment and employment in 2023

```{r}
box_simple <- svyglm(attending_college ~ sex + age + hhincome_cat,
                     family = quasibinomial,
                     design = undoc_data1,
                     na.action = na.omit)
summary(box_simple)
```


```{r}
# survey-weighted logistic regression using the svyglm() function from the survey package
# It models the probability of attending college among undocumented youth using various predictors.
box1<-svyglm(attending_college~ isrt+ driveLic+ sex+age+birthplace1+daca_imm+
                      hhincome_cat+state_enroll23+state_employ23+
                      box_enroll23+box_employ23,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit)
summary(box1)
```
 
## Model 2: logit 1b with employment and enrollment rates for 2023 (enrollment is categorical)


```{r logit1b with enroll23 and employ23}

logit1b<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+birthplace1+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit1b)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),
         p.value=round(p.value,4))
```


### Wald test for logit1b (using 2023 employment and enrollment data)
```{r wald tests for logit1}

wald_test_full_logit1_23 <- regTermTest(logit1b, ~isrt+driveLic+
                      sex+age+birthplace2+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23)
print(wald_test_full_logit1_23)

# wald test for isrt for logit1

wald_test_isrtlog1<-regTermTest(logit1b, ~isrt)
print(wald_test_isrtlog1) ## significant nice
```


### Predicted probs for logit1b

```{r logit1b bpl}
#visualize the predicted probabilities of college attendance based on birthplace1 from logit1b model. 
log1b_pred_bpl<-ggpredict(logit1b, terms="birthplace1")
print(as.data.frame(log1b_pred_bpl))
log1b_pred_bpl
plot(log1b_pred_bpl)
```

```{r logit1b isrt}
log1b_pred_isrt<-ggpredict(logit1b, terms="isrt")
print(as.data.frame(log1b_pred_isrt))
log1b_pred_isrt
plot(log1b_pred_isrt)
```




## logit2: driver license and Isrt interaction  


```{r}
logit2<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+birthplace1+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23+isrt:driveLic,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit2)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),
         p.value=round(p.value,4)) 
```



## logit 3: An employed and ISRT interaction
```{r}
logit3<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+birthplace1+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23+isrt:employ,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit3)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),
         p.value=round(p.value,4)) ## almost significant--interesting
```

### predicted probabilities for logit3--employed X isrt interaction 

```{r}
logit3_pred<-ggpredict(logit3, terms=c("isrt","employ"))
print(as.data.frame(logit3_pred))
logit3_pred
plot(logit3_pred)
```

## logit 4: birthplace and isrt interaction 

```{r}
logit4<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+birthplace1+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23+isrt:birthplace1,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit4)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),p.value=round(p.value,4)) 
```
### predicted probabilities for isrt X birthplace interaction
```{r}
logit4_pred<-ggpredict(logit4, terms=c("birthplace1","isrt"))
logit4_pred
print(as.data.frame(logit4_pred))
plot(logit4_pred)
```

## Logit 5: Race instead of birthplace 

```{r}
logit5<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+race3+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit5)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),p.value=round(p.value,4)) # isrt isn't significant

```  


## Logit6: ISRT X state employment rate interaction  

```{r}
logit6<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+race3+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23+state_enroll23_cat:isrt,
                    family=quasibinomial,
                    design=undoc_data1, 
                    na.action = na.omit))
# tidy summary
tidy(logit6)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),p.value=round(p.value,4)) # isrt isn't significant
```
### predicted probs for logit6 (isrt X state enroll interaction)
```{r}
log6_pred<-ggpredict(logit6, terms=c("state_enroll23_cat", "isrt"))
log6_pred
print(as.data.frame(log6_pred))
plot(log6_pred)
```

# Logistic Regressions with the Latinx sub-set sample  

## log_latx1 with region birthplace 3 and 2023 employment and enrollment data

```{r}

log_latx1<-(svyglm(attending_college~isrt+driveLic+
                      sex+age+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23_cat+state_employ23_cat,
                    family=quasibinomial,
                    design=lat_sub1, 
                    na.action = na.omit))

tidy(log_latx1)%>% 
  mutate(estimate=round(estimate,2),
         std.error=round(std.error,2),
         statistic=round(statistic,2),
         or=round(exp(estimate),2),
         p.value=round(p.value,4))
```

### Wald Test for effect of ISRT with Latinx sub-sample

```{r wald tests for log_latx1}

wald_full_latx1 <- regTermTest(log_latx1, ~isrt+driveLic+
                      sex+age+daca_imm+
                      employ+hhincome_cat+
                      state_enroll23+state_employ23)
print(wald_full_latx1)

# wald test for isrt for logit1

wald_isrt_latx1<-regTermTest(log_latx1, ~isrt)
print(wald_isrt_latx1) ## significant nice
```
### predicted probabilities for log_latx1
```{r}
log_latx1_pred1<-ggpredict(log_latx1, terms=c("isrt"))
log_latx1_pred1
print(as.data.frame(log_latx1_pred1))
plot(log_latx1_pred1)
```



```{r}
# the effect of DACA status on college attendance differs by household income level.
logit_daca_income <- svyglm(
  attending_college ~ daca_imm * hhincome_cat + 
    sex + age + isrt + driveLic + employ + 
    state_enroll23_cat + state_employ23_cat,
  design = undoc_data1,
  family = quasibinomial,
  na.action = na.omit
)
summary(logit_daca_income)
```



```{r}
undoc_data_18_21 <- subset(undoc_data1, age >= 18 & age <= 21)
logit_age1821 <- svyglm(
  attending_college ~ isrt + driveLic + sex + age + race3 + daca_imm +
  employ + hhincome_cat + state_enroll23_cat + state_employ23_cat,
  design = undoc_data_18_21,
  family = quasibinomial
)
summary(logit_age1821)
```

```{r}
lat_sub1_age1821 <- subset(lat_sub1, age >= 18 & age <= 21)
log_latx1_age1821 <- svyglm(
  attending_college ~ isrt + driveLic + sex + age + daca_imm +
  employ + hhincome_cat + state_enroll23_cat + state_employ23_cat,
  design = lat_sub1_age1821,
  family = quasibinomial
)
summary(log_latx1_age1821)
```



